---
title: "G10 FX Returns: A Statistical Analysis"
author: "NavyBlueCheese"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 6,
  dpi = 150
)
```

## Overview

This notebook pulls two years of daily G10 FX spot rates from the
[Frankfurter API](https://frankfurter.dev), computes daily log returns, and
runs a set of classic quantitative checks:

- Pairwise **correlation structure** across all nine currency pairs
- **Return distributions** compared against a Normal fit (fat tails?)
- **Rolling 30-day Sharpe ratios** to track risk-adjusted performance over time
- **Volatility clustering** as a motivating observation for GARCH-type models

All nine G10 currencies (EUR, JPY, GBP, AUD, NZD, CAD, CHF, NOK, SEK) are
quoted against USD throughout.

---

## 1. Libraries

```{r libraries}
library(jsonlite)
library(dplyr)
library(tidyr)
library(ggplot2)
library(purrr)
library(lubridate)
library(patchwork)
library(scales)
library(zoo)
```

---

## 2. Fetch Data

The Frankfurter API returns USD-based rates for all G10 currencies in a single
request. One call covers the entire two-year window.

```{r fetch, cache=TRUE}
G10 <- c("EUR", "JPY", "GBP", "AUD", "NZD", "CAD", "CHF", "NOK", "SEK")

end_date   <- Sys.Date() - 1
start_date <- end_date - years(2)

fetch_rates <- function(start, end, base = "USD") {
  url <- sprintf(
    "https://api.frankfurter.dev/v1/%s..%s?from=%s",
    format(start, "%Y-%m-%d"),
    format(end,   "%Y-%m-%d"),
    base
  )
  # simplifyDataFrame=FALSE keeps rates as a named list of named vectors
  # (keys = dates) rather than letting jsonlite collapse it into a data.frame
  # where names() would return currency codes instead of date strings.
  resp <- fromJSON(url, simplifyDataFrame = FALSE)
  map_dfr(names(resp$rates), function(d) {
    r <- resp$rates[[d]]
    tibble(date = as.Date(d), currency = names(r), rate = as.numeric(unlist(r)))
  }) %>%
    filter(currency %in% G10)
}

fx <- tryCatch(
  fetch_rates(start_date, end_date),
  error = function(e) {
    stop("API fetch failed: ", e$message,
         "\nCheck your connection or try again later.")
  }
)

cat(sprintf("Trading days: %d | Date range: %s to %s\n",
            n_distinct(fx$date),
            min(fx$date), max(fx$date)))
```

---

## 3. Log Returns and Summary Statistics

Daily log returns are computed as `log(rate_t / rate_{t-1})`. Annualised
figures use 252 trading days.

```{r returns}
fx_ret <- fx %>%
  arrange(currency, date) %>%
  group_by(currency) %>%
  mutate(log_ret = log(rate / lag(rate))) %>%
  ungroup() %>%
  drop_na(log_ret)

stats <- fx_ret %>%
  group_by(currency) %>%
  summarise(
    ann_return = mean(log_ret) * 252,
    ann_vol    = sd(log_ret) * sqrt(252),
    sharpe     = ann_return / ann_vol,
    skewness   = mean((log_ret - mean(log_ret))^3) / sd(log_ret)^3,
    ex_kurt    = mean((log_ret - mean(log_ret))^4) / sd(log_ret)^4 - 3,
    n_obs      = n(),
    .groups    = "drop"
  ) %>%
  arrange(desc(sharpe))

knitr::kable(
  stats %>% mutate(across(ann_return:ex_kurt, ~ round(.x, 4))),
  col.names = c("Currency", "Ann. Return", "Ann. Vol", "Sharpe",
                "Skewness", "Ex. Kurtosis", "N"),
  caption   = "Table 1: G10 FX summary statistics vs USD (2-year period)"
)
```

---

## 4. Correlation Heatmap

```{r correlation, fig.height=7}
ret_wide <- fx_ret %>%
  select(date, currency, log_ret) %>%
  pivot_wider(names_from = currency, values_from = log_ret) %>%
  select(-date) %>%
  drop_na()

corr_mat  <- cor(ret_wide)
corr_long <- as.data.frame(corr_mat) %>%
  tibble::rownames_to_column("x") %>%
  pivot_longer(-x, names_to = "y", values_to = "corr")

# Order currencies by a simple hierarchical clustering for visual clarity
ord <- hclust(as.dist(1 - corr_mat))$order
lvls <- rownames(corr_mat)[ord]

corr_long <- corr_long %>%
  mutate(
    x = factor(x, levels = lvls),
    y = factor(y, levels = lvls)
  )

ggplot(corr_long, aes(x, y, fill = corr)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = round(corr, 2)), size = 3.3,
            color = ifelse(abs(corr_long$corr) > 0.6, "white", "black")) +
  scale_fill_gradient2(
    low      = "#ef4444",
    mid      = "#f8fafc",
    high     = "#14b8a6",
    midpoint = 0,
    limits   = c(-1, 1),
    name     = "Pearson r"
  ) +
  labs(
    title    = "G10 FX Return Correlations (vs USD)",
    subtitle = "Daily log returns, past 2 years | Rows/columns ordered by hierarchical clustering",
    x = NULL, y = NULL,
    caption  = "Source: frankfurter.dev"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text      = element_text(size = 11, face = "bold"),
    plot.title     = element_text(size = 14, face = "bold"),
    plot.subtitle  = element_text(size = 10, color = "#64748b"),
    legend.position = "right",
    panel.grid     = element_blank()
  )
```

---

## 5. Return Distributions

Normal returns would hug the dashed red curve. FX returns almost always show
heavier tails (positive excess kurtosis) — meaning extreme moves are more
frequent than a Gaussian model would predict, with direct implications for
Value-at-Risk estimates.

```{r distributions, fig.height=9}
normal_curves <- fx_ret %>%
  group_by(currency) %>%
  summarise(
    mu    = mean(log_ret * 100),
    sigma = sd(log_ret * 100),
    .groups = "drop"
  ) %>%
  pmap_dfr(function(currency, mu, sigma) {
    x_seq <- seq(mu - 4.5 * sigma, mu + 4.5 * sigma, length.out = 400)
    tibble(currency = currency, x = x_seq, y = dnorm(x_seq, mu, sigma))
  })

ggplot(fx_ret, aes(x = log_ret * 100)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins  = 70,
    fill  = "#14b8a6",
    color = "white",
    alpha = 0.75
  ) +
  geom_line(
    data = normal_curves,
    aes(x = x, y = y),
    color     = "#ef4444",
    linewidth = 0.8,
    linetype  = "dashed"
  ) +
  facet_wrap(~currency, ncol = 3, scales = "free") +
  labs(
    title    = "Daily Log Return Distributions",
    subtitle = "Teal bars = empirical density | Red dashed = Normal fit",
    x        = "Daily Log Return (%)",
    y        = "Density",
    caption  = "Source: frankfurter.dev"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text    = element_text(face = "bold"),
    plot.title    = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10, color = "#64748b")
  )
```

---

## 6. Rolling 30-Day Sharpe Ratio

A 30-day rolling Sharpe (annualised) captures how quickly the risk-adjusted
return regime shifts. The dotted lines at +/-1 indicate a commonly used
threshold for "decent" risk-adjusted returns.

```{r rolling-sharpe, fig.height=9}
roll_sharpe <- fx_ret %>%
  arrange(currency, date) %>%
  group_by(currency) %>%
  mutate(
    r_mean   = rollapply(log_ret, 30, mean, fill = NA, align = "right"),
    r_sd     = rollapply(log_ret, 30, sd,   fill = NA, align = "right"),
    r_sharpe = (r_mean * 252) / (r_sd * sqrt(252))
  ) %>%
  ungroup() %>%
  drop_na(r_sharpe)

ggplot(roll_sharpe, aes(x = date, y = r_sharpe, color = currency)) +
  geom_hline(yintercept = c(-1, 1), linetype = "dotted",
             color = "#cbd5e1", linewidth = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed",
             color = "#94a3b8", linewidth = 0.5) +
  geom_line(alpha = 0.85, linewidth = 0.55) +
  facet_wrap(~currency, ncol = 3) +
  scale_color_brewer(palette = "Set2") +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %y") +
  labs(
    title    = "Rolling 30-Day Sharpe Ratio (annualised)",
    subtitle = "vs USD | Dashed = 0 | Dotted = +/- 1",
    x        = NULL,
    y        = "Sharpe Ratio",
    caption  = "Source: frankfurter.dev"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "none",
    strip.text      = element_text(face = "bold"),
    plot.title      = element_text(size = 14, face = "bold"),
    plot.subtitle   = element_text(size = 10, color = "#64748b"),
    axis.text.x     = element_text(size = 8, angle = 30, hjust = 1)
  )
```

---

## 7. Volatility Clustering

Absolute returns proxy realised volatility. If volatility clusters (a GARCH
stylised fact), spikes should appear in bursts rather than uniformly across
time. The LOESS trend line makes this visible.

```{r vol-clustering, fig.height=9}
fx_ret %>%
  mutate(abs_ret = abs(log_ret) * 100) %>%
  ggplot(aes(x = date, y = abs_ret)) +
  geom_line(color = "#14b8a6", alpha = 0.5, linewidth = 0.35) +
  geom_smooth(
    method = "loess", span = 0.12, se = FALSE,
    color = "#0f172a", linewidth = 0.9
  ) +
  facet_wrap(~currency, ncol = 3, scales = "free_y") +
  labs(
    title    = "Absolute Daily Log Returns (Realised Volatility Proxy)",
    subtitle = "Teal = daily | Dark = LOESS trend (span = 0.12)",
    x        = NULL,
    y        = "|Log Return| (%)",
    caption  = "Source: frankfurter.dev"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text    = element_text(face = "bold"),
    plot.title    = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10, color = "#64748b"),
    axis.text.x   = element_text(size = 8)
  )
```

---

## 8. Key Takeaways

- **Fat tails confirmed**: Every currency shows positive excess kurtosis, so
  a Normal VaR model will systematically underestimate tail risk.

- **Bloc correlations**: EUR, NOK, SEK, and CHF cluster together (European
  risk bloc). AUD and NZD are highly correlated with each other (commodity
  currencies). JPY and CHF are frequently negatively correlated with
  risk-on currencies during stress.

- **Sharpe regimes are short-lived**: No currency holds a consistently high
  or low Sharpe vs USD across the full two-year window. Regime changes
  dominate any carry of the "trend."

- **Volatility clustering visible**: The LOESS trend clearly shows periods of
  elevated realised vol, consistent with GARCH(1,1) dynamics. This motivates
  using time-varying volatility models rather than constant-vol assumptions.

---

*Data: [Frankfurter API](https://frankfurter.dev) — free, no API key required.
For educational purposes only.*
